### 2016-09-21 doc/publications/plos16/plos16.tex
changes to table in plos16.tex

Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
The joint interaction is achieved by appling the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example, by summing the number of times a variable is used to split a node, over all nodes and all trees.
RF are well suited for processing ``wide'' genomic data for two reasons. 
Firstly, while other machine learning applications have the propensity to overfit datasets with more features $p$ than samples $n$ (``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}), RF are known to be more resistant through not immune to overfitting~\cite{Segal.2004}.
Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow separate trees on each processor and combine the mature trees.

The most popular random forest software packages are R-based, although current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}. 
This is despite R now supporting long vectors (length greater than $2^{31}$), and the standard {\sc random forest} implementation in R using paralelization to grow multiple trees simultaneously~\cite{Liaw2002}. 
%While R has provisions to support long vectors (length greater than $2^{31}$), current R-based RF implementations, such as {\sc random forest} or {\sc party}~\cite{Hothorn2006}, do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  
{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by these methods. 
The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) as oppose to using virtually infinite numbers of CPUs across multiple nodes. This in turn limits the number of tasks that can be performed in parallel such that only splitting at whole tree-level is feasible as oppose to a more fine grained strategy of handing off the calculation for each tree node to a separate processor. 
Hadoop/Spark overcomes these limitation by enabling programs to scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for ``big" (many samples) and ``wide" (many features) data sets. \cursedforest\ is capable of parallelising the split for each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small datasets where multi-threading is considered more performant. 
\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets. 
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis, VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. 
This offers a comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions in older array technology~\cite{Howie2012}. 
This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.

%Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.  
%Therefore, these historically successful solutions are not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, classification or feature importance analysis needs to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
%For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
%This dataset consumes close to a terabyte of disk space.  
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full
%$2500 \times 80$ million matrix to cluster individuals by
%their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
%Although Spark ML can build a RF model on
%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
%Spark ML uses, which we elaborate on in the discussion.
%
%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
%Unlike
%the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
%excessive resource requirements on the driver or worker nodes.
%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
can have on the prediction accuracy of a model.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
% >>CONDENSE
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\subsection{Overview of Random Forests}

@@

% <<CONDENSE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the cursed forest implementation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cursed Forest.}
% DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
similar to DataFrames in R.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
\subsubsection{1000 Genomes}
%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Results
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature selection 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest grows denser forests to stabilise feature importance ranking}
In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.

To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DENIS: Rob I don't think this is relevant anymore?
% >>CONDENSE
%\marginpar{These are regression examples, we need some classification examples}
%To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
%much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
%predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
%the training data set from $n=100 \times p=1000$ to $p=1000000$.
%%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
%%TODO also how is the added noise playing into this ?
%The random forest model recovers variables 1 and 4 as the two
%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

%% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
%% can recover the variables defining a functional relationship in the presence of a substantial number of noise
%% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
%% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
%% to be altered from their default value when there are a large number of predictors.

%% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
%% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
%% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
%% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
%% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
%% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

%% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
%% the significant varaiables if the functional relationship is simple.

%% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
%% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?

%\begin{figure}[tbhp]
%\begin{tabular}{ll}
%a)& b)\\
%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
% \end{tabular}
% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
%   variable with added Gaussian noise (in red).}
%  \label{figure:synth}
%\end{figure}
% <<CONDENSE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Feature selection application example 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Feature selection application example}
%Maybe we should have a feature selection on real data section. We could be using one of the smaller datasets used in the other papers?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest outperforms existing methods for classification problems}
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the 1000 genomes dataset.

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
\hline
%\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  &  &  \\
&&& RF - Spark ML &  &  &  \\
&&& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
&&& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
 \hline
phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
&&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
&&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
\hline
phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
&&& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
&&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
&&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
  was done on dpas-03} \\ 
\hline
phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
&&& RF - Spark ML & &  &\\
&&& LR - Spark ML & & 14hr 48min & 8GB \\ 
&&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
\hline
phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
&&& RF - Spark ML & - & - & - \\
&&& LR - Spark ML & - & - & - \\ 
%&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. 
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Sensitivity 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sensitivity of parameter choice}

@@

\section{Conclusion}

@@

@@

%  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
%    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
%  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
%    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

%@article{Chen2012323,
%title = "Random forests for genomic data analysis ",
%journal = "Genomics ",
%volume = "99",
%number = "6",
%pages = "323 - 329",
%year = "2012",
%note = "",
%issn = "0888-7543",
%doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
%url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
%author = "Xi Chen and Hemant Ishwaran",
%keywords = "Random forests",
%keywords = "Random survival forests",
%keywords = "Classification",
%keywords = "Prediction",
%keywords = "Variable selection",
%keywords = "Genomic data analysis ",
%abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features. This makes \{RF\} particularly appealing for high-dimensional genomic data analysis. In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning. "
%}

>>>>

Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
The joint interaction is achieved by apply the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided for example by summing the number of times a variable is used to split a node, over all nodes and all trees.
Processing ``wide'' genomic data poses difficulties for machine learning applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalise poorly. 
This is known as the ``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}. 
RF are known to be more resistant to overfitting than other techniques~\cite{Segal.2004}. 
%DENIS Rob, Piotr anything else to add or explain more clearly?
\marginpar{Rob, Piotr anything else to add or explain more clearly?}
RF are hence well suited for modelling complex polygenic diseases and identifying the strongest contributing features overall.

RF are very easy to parallize. As the forest is a sum of decision trees, it it possible to grow separate trees on 
each processor and sum them. This mode of parallization is widely used for the {\sc random forest} implementation in
R~\cite{}, and could also be used for the {\sc party}~\cite{} package.
\marginpar{Rob, Piotr can you extend this with some more detailed descriptions of party and RF in R}
 However,
this has several limitations. One is that these implementations reach a hard limit in R when they require vectors of a
length greater than $2^{31}-1$. While R now has basic support for long vectors, not all functions can utilize it.
These traditional implementations are have been reported to not scale well with increasing feature
size~\cite{Wright.and.Ziegle.2016}.

However a more basic problem may be that the parallization is at too course a grain. The Cursed Forest implementation has
a finer grained version of parallization where the calculation for each split is handed off to a separate processor.

Historically successful solutions are hence not catering for the big data demands that especially due to WGS are becoming more prevalent. Here, GWAs-type analyses need to be performed on datasets with a much larger number of features~\cite{Yano2016}. 
For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
This dataset consumes close to a terabyte of disk space.  
The richness of this data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions e.g. from older older array technology~\cite{Howie2012}.
This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.

%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.  
%DENIS Rob do you want to flesh this out more
\marginpar{Rob do you want to flesh this out more}
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.

However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features. 
Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
Using VariantSpark, we successfully built a k-means model on the full
$2500 \times 80$ million matrix to cluster individuals by
their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

Here we extend this toolkit to encompass supervised learning using random forest. 
Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
Although Spark ML can build a RF model on
a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
Spark ML uses, which we elaborate on in the discussion.

We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
% Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
% excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
% the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do
  we need to spell this out better, it should be a result section.}  
Harnessing the virtually unlimited capability to parallelise
tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc
  r2VIM}~\cite{Szymczak2016} on smaller datasets.

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
can have on the prediction accuracy of a model.

\section{Methods}

% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\section{Overview of Random Forests}

@@

\subsection{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
similar to DataFrames in R.

@@

\section{Results and Discussion}

\marginpar{These are regression examples, we need some classification examples}

To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
the training data set from $n=100 \times p=1000$ to $p=1000000$.
%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
%TODO also how is the added noise playing into this ?
The random forest model recovers variables 1 and 4 as the two
most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
% can recover the variables defining a functional relationship in the presence of a substantial number of noise
% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
% to be altered from their default value when there are a large number of predictors.

% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
% the significant varaiables if the functional relationship is simple.

% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?

\begin{figure}[tbhp]
\begin{tabular}{ll}
a)& b)\\
 \includegraphics[totalheight=6cm]{./figs/tree_function.png}&
  \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
 \end{tabular}
 \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
   variable with added Gaussian noise (in red).}
  \label{figure:synth}
\end{figure}

\subsection{CursedForest processes genome-wide variant data}
%\label{comp}
%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
%time/mem)

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}                               \\ \hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest  & 0.077 (XX) &                  &                                                                   \\
                                                & RF - Spark ML &            &                  &                                                                   \\
                                                & RF - R        & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger -R     & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger c++    & 0.94       & 7 mins           & --nthreads 20                                                     \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                                     \\
                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)}                \\
                                                    & ranger  -R    & 0.94       & 4 1/2 mins       & 21GB (one processor)                                                                \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & CursedForest  & 0.063 (XX) & 1hr 58min        & 4GB                                                                                 \\
                                                    & RF - Spark ML & 0.94       & 8hr 29min        & 8GB                                                                                 \\
                                                    & LR - Spark ML &            & 2hr 6min         & 8GB                                                                                 \\ 
                                                    & ranger  R     & 0.95       & 28 mins          & num.threads=10\footnote{don't
                                                                                                      know how to get the memory
                                                                                                      used,   This was done on dpas-03}  \\ 
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & CursedForest  & 0.046 (XX) & 4hr 0min                                             & 4GB         \\
                                                    & RF - Spark ML &            &                                                      &             \\
                                                    & LR - Spark ML &            & 14hr 48min                                           & 8GB         \\ 
                                                    & ranger  R     &  0.97      & 58 mins\footnote{ but took 6 hours to load the
                                                                                   data}
                                                                                                                                         &
                                                                                                                                           909GB\footnote{I
                                                                                                                                           think
                                                                                                                                           this
                                                                                                                                           is
                                                                                                                                           total
                                                                                                                                           memory. The
                                                                                                                                           job
                                                                                                                                           was
                                                                                                                                           run
                                                                                                                                           with
                                                                                                                                                   ntasks-per-node=160
                                                                                                                                           cores-per-socket=10
                                                                                                                                           on
                                                                                                                                           ruby
                                                                                                                                           }  \\ 
& ranger   -R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
& ranger (C++)       &        NA     &        NA     &             80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA      &            160 CPU -- runs out of time at 20 hours \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. Size of the data is listed as individuals x variants.

\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

%\begin{table}
%\begin{minipage}{\textwidth}
%\begin{center}
%\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
%\hline
%\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
%\hline
%RF (R)\footnote{10 nodes, 500 trees, dpas03} 
 %                                  &  0.456 & 2.89 & 18GB \\ \hline
%CursedForest\footnote{something}
 %                                  & X & X & X \\ \hline
%\end{tabular}
%\medskip
%\begin{flushleft} 
%\caption{Comparion between Random Forests on R and CursedForests on a smaller
%dataset. Chromosome 22 - phase1 (2.1GB) }
%\label{table:smaller_data}
%\end{flushleft}
%\end{center}
%\end{sidewaystable}
%\end{minipage}
%\end{table}

% \begin
% \begin{table}[!ht]
% \caption{
% {\bf Performance comparison between the different Machine Learning algorithms.}}
% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
% \hline
% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
% \hline
%  chr 	 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
% \end{tabular}
% \begin{flushleft} 
% \end{flushleft}
% \label{table1}
% %\end{adjustwidth}
% \end{table}

@@

We note that it is possible for a RF model to have a very poor fit, and be useless for prediction, and still have a 
useful ranking of variable importance.

@@

%\subsection{James' dataset}
%Comparing CursedForest with feature selection + linear model (performance and time/mem)

\subsection{Conclusion}

@@

%\section*{References}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
%

@@

  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

@article{Chen2012323,
title = "Random forests for genomic data analysis ",
journal = "Genomics ",
volume = "99",
number = "6",
pages = "323 - 329",
year = "2012",
note = "",
issn = "0888-7543",
doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
author = "Xi Chen and Hemant Ishwaran",
keywords = "Random forests",
keywords = "Random survival forests",
keywords = "Classification",
keywords = "Prediction",
keywords = "Variable selection",
keywords = "Genomic data analysis ",
abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features. This makes \{RF\} particularly appealing for high-dimensional genomic data analysis. In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning. "
}
